Implementing simplified INLA into `aghq`

### 2022-10-21 src/dev_sinla/sinla.Rmd
Normalise the Laplace with aghq and map argument to MakeADFun

Let's have a go at computing Equation \@ref(eq:pxigiveny) so that we can compare to the posterior distribution from `tmbstan`.
We simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to:
$$
\begin{align}
\tilde p(x_i, \hat \theta \, | \, y) &= \tilde p_\text{LA}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
&\propto \tilde p_\text{LA}(x_i \, | \, \hat \theta, y)
\end{align}
$$

Start by getting $\hat \theta$:

@@

#' How many equally spaced points do we want to evaluate the posterior at?
spacing <- 0.05
x_grid <- seq(from = 0.5, to = 3.5, by = spacing)
x_grid_length <- length(x_grid)

df <- opt$par %>%
  t() %>%
  as.data.frame() %>%
  slice(rep(1:n(), each = x_grid_length)) %>%
  mutate(
    phi_prev_i = x_grid
  ) %>%
  rowwise() %>%
  mutate(
    #' Note here: it's crucial that the arguments are in the right order!
    logpost = -obj$fn(c(beta_prev, phi_prev_i, log_sigma_phi_prev))
  )
```

We've compuating the unnormalised log-posterior -- how can we go from that to a normalised posterior?
Going to do trapezoid rule here, but AGHQ would work better (even with a low number of points) because the function is close to being Gaussian.

### Safe exponential then trapezoid rule {-}

@@

### `logSumExp` to calculate the log normalising constant {-}

@@

### Calculating $p_\text{LA}(\hat \theta \, | \, y)$ with `aghq`

```{r}
h$he <- function(theta) numDeriv::jacobian(h$gr, theta, method = "Richardson")
quad <- aghq::aghq(ff = h, k = 3, startingvalue = c(0, 0), control = aghq::default_control_tmb())
quad$normalized_posterior$lognormconst
```

@@

>>>>

Let's have a go at computing Equation \@ref(eq:pxigiveny) so that we can compare to the posterior distribution from `tmbstan`.
We simplify matters by considering just one quadrature point, $\hat \theta$, such that the sum over $z$ collapses to^[Note that I'm uncertain if $\hat \theta$ (a constant) should still be included in the LHS of this.]:
$$
\begin{align}
\tilde p(x_i \, | \, y) &= \tilde p_\text{LA}(x_i \, | \, \hat \theta, y) \tilde p_\text{LA}(\hat \theta \, | \, y) \\
&\propto \tilde p_\text{LA}(x_i \, | \, \hat \theta, y)
\end{align}
$$

We're always going to need $\hat \theta$.
We have calculated it before, but we will get it fresh here (because of questionable programming practices whereby objects are overwritten):

@@

opt$par
```

Now, a basic way we could approach this is to define a grid of points that we would like to evaluate the posterior at.
We can then create a dataframe where each row contains the values `opt$par` for $\theta$ and a grid value for $x_i$.
Each row can then be passed into `-obj$fn()` to calculate the log-posterior.
Note that it's crucial the arguments are in the right order when we pass them to `obj$fn()` -- this has caused headaches!

```{r}
spacing <- 0.05
x_grid <- seq(from = 0.5, to = 3.5, by = spacing)
x_grid_length <- length(x_grid)

df <- t(opt$par) %>%
  as.data.frame() %>%
  slice(rep(1:n(), each = x_grid_length)) %>%
  mutate(phi_prev_i = x_grid) %>%
  rowwise() %>%
  mutate(logpost = -obj$fn(c(beta_prev, phi_prev_i, log_sigma_phi_prev)))
```

We've computed the unnormalised log-posterior -- how can we go from that to a normalised posterior?
Given these grid of points, we can do the trapezoid rule here.

### Safe exponential trapezoid rule {-}

@@

### `logSumExp` trapezoid rule {-}

@@

## Normalisation with `aghq`

@@

So how might we approach this in a more sophisticated way?
We could do the normalisation directly in `aghq`!

We could normalise $\tilde p_\text{LA}(\theta, y)$.
Note that sometimes we have been calling this $\tilde p_\text{LA}(\theta \, | \, y)$ because they are equivalent up to a constant of proportionality.
Define
$$
\tilde p_\text{AQ}(\theta \, | \, y) = \frac{\tilde p_\text{LA}(\theta, y)}{\tilde p_\text{AQ}(y)}
$$
or on a log-scale
$$
\log \tilde p_\text{AQ}(\theta \, | \, y) = \log \tilde p_\text{LA}(\theta, y) - \log \tilde p_\text{AQ}(y)
$$

First we have to add a numerical Hessian that's missing from `h`:

```{r}
h$he <- function(theta) numDeriv::jacobian(h$gr, theta, method = "Richardson")
```

Now we can obtain $\log \tilde p_\text{AQ}(y)$:

```{r}
quad <- aghq::aghq(ff = h, k = 3, startingvalue = c(0, 0), control = aghq::default_control_tmb())
quad$normalized_posterior$lognormconst
```

Remember that `h` is the negative log-likelihood, so we want to add the normalising constant rather than subtract it:

```{r}
h_norm <- function(theta) h$fn(theta) - quad$normalized_posterior$lognormconst
```

Now we can put this into
$$
\tilde p(x_i, \hat \theta \, | \, y) = \tilde p_\text{LA}(x_i \, | \, \hat \theta, y) \tilde p_\text{AQ}(\hat \theta \, | \, y)
$$

But perhaps there is no reason to, because we will just have to normalise it again, so why do it twice?
Let's just try to normalise it directly.
What we need to do is create a function in `TMB` where we fix some of the inputs to the function.
This can be done using the `map` argument of `MakeADFun`.
You set each parameter to `factor(NA)` that you want to be fixed, then they will retain the value you input in `parameters`:

```{r}
obj$par %>% names()

map_fixed_theta <- list(beta_prev = factor(NA), log_sigma_phi_prev = factor(NA))

param_i_fixed_theta <- param_i

theta_names <- names(opt$par)
for(theta in theta_names) {
  param_i_fixed_theta[[theta]] <- opt$par[[theta]]
}

dat_i <- prepare_dat(dat, i)

obj_fixed_theta <- MakeADFun(
  data = dat_i,
  parameters = param_i_fixed_theta,
  random = "phi_prev_minus_i",
  DLL = "model1index",
  silent = TRUE,
  map = map_fixed_theta
)
```

Now we can normalise this function $p_\text{LA}(x_i, \theta = \hat \theta, y)$ using `aghq`:

```{r}
quad <- aghq::aghq(ff = obj_fixed_theta, k = 3, startingvalue = 0, control = aghq::default_control_tmb())
```

The value of the normalising constant is (approximately) the same as we get using our trapezoid rule:

```{r}
abs(quad$normalized_posterior$lognormconst - norm_trapezoid_log) < 10E-3
```
